\thetitle

### 2013-04-19 tfpie/tfpie.tex
progress on 4

Could work with any fashion, and should really lead into many
fashionable languages.  The purpose of showing many is to demonstrate
that the same principles apply in each setting.  Each language can also
expose students to various mechanisms.

@@

class Examples {
    void test_tree(Tester t) {
        t.checkExpect(new Leaf(7).sum(), 7);
        t.checkExpect(new Node(new Leaf(1),
                               5, 
                               new Node(new Leaf(0), 10, new Leaf(0))).sum(),
                      16);
    }
}
\end{verbatim}
\caption{Binary tree sum in How to Design Classes}
\end{figure}

\subsection{Java}

\begin{itemize}
\item Types
\item Compiler
\item ``Industrial'' features (collections library, ...)
\end{itemize}

\subsection{Unifying functions and objects}

>>>>

The transition from custom teaching languages to a professional
language takes place about half-way through the course.  At this
point, students have already seen many of the essentials of
object-oriented programming. In particular: object, classes, fields
and methods, dynamic dispatch, inheritance, and overriding.

From this point, almost any language that students might use in future
co-op positions and courses would be an appropriate follow-up.  Our
course transitions to Java, but C\#, Python, Ruby, Eiffel, or
JavaScript would all work naturally.  The key lesson of the transition
is that the fundamental principles underlying object-oriented
programming remain the same between languages, and that learning a new
language is primarily a matter of mapping these concepts to specific
constructs in the new language.  Of course, particular languages also
use unique specific mechanisms which need to be taught to use the
language effectively, but these are rarely as vital as the
cross-language principles.

\subsection{Functional Java}

The transition begins with replicating the object-oriented style of
our teaching languages in Java.  In particular, we do not introduce
mutation, for loops, or mutable data structures such as arrays or
\texttt{ArrayList}s until later in the semester.  Instead, students
design data representations using classes, with \texttt{interface}s
representing unions of data.  Additionally, we avoid mention of the
distinction between primitive and other values in Java, which is made
easier by not using standard libraries early.  An example of this
style of programming is presented in figure~\ref{fig:java}, repeating
the binary tree sum from the previous section.

@@

class Examples {
    void test_tree(Tester t) {
        t.checkExpect(new Leaf(7).sum(), 7);
        t.checkExpect(new Node(new Leaf(1),
                               5, 
                               new Node(new Leaf(0), 10, new Leaf(0))).sum(),
                      16);
    }
}
\end{verbatim}
\caption{Binary tree sum in How to Design Classes}
\label{fig:java}
\end{figure}

Comparing this figure to the previous example illustrates a number of the
differences that students are exposed to upon transition to Java.

\begin{enumerate}
\item Explicit representation of unions and interfaces in the
  language.  Previously, interfaces were simply described in stylized
  comments, following the \htdp approach.
\item Types are now specified as part of the program, and are now
  enforced.
\item Java syntax is substantially different and more verbose. For
  example, constructors must be defined explicitly.
\item The testing environment is somewhat different, and requires
  additional boilerplate, although we are
  able to use the JavaLib framework~\cite{local:java-world} to support simple testing by
  structural equality.
\end{enumerate}

\noindent
Of course, there are other differences which cannot be seen from a
code snippet.

\begin{enumerate}
  \setcounter{enumi}{4}
\item Students must use a new development environment and compiler.
  In class, we primarily develop in a text editor and run the Java
  compiler at the command line.  In labs and on homeworks, students
  typically use the Eclipse IDE.
\item Installing and configuring libraries is now required.  Because
  we use a custom library for testing, students must cope with library
  installation and class paths on the first day.
\end{enumerate}

Of course, all but the first two of these changes are unrelated to the
fundamental lessons we hope to teach---the rest merely present
additional hurdles for students.

\subsection{Traditional Java}

Thanks to the the preparation in the first half of the course, we can
cover functional OO programming in Java in a just a few lectures.  We
then increase the subset of the language we use to encompass mutation,
loops, and mutable data structures. We present \texttt{ArrayList}s,
followed briefly by arrays. Students use, and then implement, hash
tables as well as other mutable and immutable data structures.
Conventional input and output are treated only very briefly, as we
focus instead of both fundamentals and exercises making use of real
APIs such as hashing functions or Twitter posting.  Finally, while,
for, and for-each loops are presented, following the methodology of
\htdc which connects loops to stylized use of recursive functions with
accumulators, a technique the students now have two semsesters of
practice with.

\subsection{Beyond Traditional Java}
